---
title: "Pie Charts"
author: "Justin Strate"
output: html_document
---
Suppose we have have a survey of 50 texans on what Texas' basketball team they prefer. Suppose we 25 say Spurs, 15 say Rockets and 10 say Mavericks. And, we record this data
in two vectors.

```{r}
x <- c(25, 15, 10)
labs  <- c('Spurs', 'Rockets', 'Mavericks')
 
#A basic pie chart
pie(x, labels = labs, main="Pie Chart of Programming")
```

Let's say we want the percentage of users in the in the label. We can do this by creating a new label(labs2).

```{r}
pcts <- paste( round(x/sum(x),3) *100, "%", sep="") #Round to 3 significant digits
labs2 <- paste(labs, pcts, sep = " ")
``` 

Let's change these bland color schemes.
```{r}
colLabs <- c('green', 'blue', 'yellow')

pie(x, labels = labs2, col=colLabs, main="Pie Chart of Programming")
```

Personally, I prefer 3d pie charts. You can use the function pie3D in the plotrix library
to make one.

```{r}
library(plotrix)
pie3D(x, labels = labs2, col=colLabs, 
       main="Pie Chart of Programming")
```

You can make the slices stick out more using the explode option.

```{r}
pie3D(x, labels = labs2, col=colLabs,  
      main="Pie Chart of Programming", explode = 0.1)
```